Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-08-11 Manuscript/MPESubmission/Revision/response.tex

% -----------------
% Comment 15
% -----------------
\item{\textcolor{blue}{\textbf{Previous research:} The missing-data problem has a long history in the literature and there are references to some of those older and recent papers.
But I feel it is necessary to summarize the following questions: what did those studies find, what are the questions those studies left open, and how does this study address them?}}

We now emphasize more of the recent findings from the missing-data papers in the introduction, and discussion (see comment 4 and 10 above).
We also included that our results for the $M_F$ parameter for the Normalised Robinson-Foulds metric were in accordance with Manos et al. 2007 in the discussion about the $M_F$ parameter:

\textit{These results are in agreement with Manos et al. (2007) where as few as 16 characters were sufficient for correctly assigning artificial fossils to their correct clade.} lines @@@

@@

\end{enumerate}

\section{Reviewer 1 Specific comments:}
\begin{enumerate}

% -----------------
% Specific Comment 1
% -----------------
\item{\textcolor{blue}{\textbf{line 31:} has it really been shown that the majority of macroevolutionary studies focuses on extant species only? The cited references are examples of large scale extant-only phylogenies. To make such a pointed statement, the authors would need to find support from some kind of literature review or meta-analysis comparing the number of paleontological to neontological macroevolutionary studies. Otherwise, rephrase. }}

We revised this statement as follows: \textit{many large-scale macroevolutionary studies focus solely on living species} (line 30-31 @@@).

@@

% -----------------
% Specific Comment 2
% -----------------
\item{\textcolor{blue}{\textbf{line 37 and throughout:} Rephrase so that 'however' does not start the sentence, such as 'To do this, however, we need…'. Sentences should not start with however if the meaning is to contrast with the previous statement (Strunk, W. and White, E.B. 2009. The Elements of Style. NY: Pearson Education. p 48-49).}}

We thank the reviewer for this stylistic suggestion and modified the placement of 'however' throughout the manuscript.

% -----------------
% Specific Comment 3
% -----------------
\item{\textcolor{blue}{\textbf{line 43:} sentence choppy, consider rephrasing or splitting into two sentences.}}

We modified the sentence as follows: \textit{These approaches differ mainly in how they treat fossil taxa and their data. One can use fossils as tips or as nodes in the phylogeny, and can use only the age of the fossils, only the morphology of the fossils, or age and morphology jointly.} (line 42-44@@@)

@@

% -----------------
% Specific Comment 11
% -----------------
\item{\textcolor{blue}{\textbf{line 82:} change to '…missing data are not phylogenetically biased, Wiens…'}}

We modified the sentence as suggested.

% -----------------
% Specific Comment 12
% -----------------
\item{\textcolor{blue}{\textbf{line 99:} unclear - what is meant by 'with a fixed topology approach'? The studies references investigated divergence times using total evidence datasets and in some cases used constraints but not completely fixed topologies}}

We modified the statement to: \textit{but also because previous studies have already empirically assessed the effect of the Total Evidence method on branch length variation but using topological constraints} (line 111-113@@@).

@@

(1) - We thank the reviewer for this comment and now emphasize both the novel aspects of our study and the questions we are actually testing (see reviewer 1 comment 4 above).

(2) - the second point underlined by the reviewer in this specific comment strikes use as really interesting.
We ran this additional analysis by comparing the estimated $\alpha$ parameter for the morphological rates distribution to the ``true'' $\alpha$ parameter.
We found results that the $\alpha$ parameter is usually slightly underestimated for the various ``missing-data'' trees unless the number of living species with no morphological data is high ($M_L$ = 75\%), then, the $\alpha$ are overestimated.
These results are somewhat in agreement with our topological differences results: the number of living species seems to be a driving missing data parameter for inferring Total Evidence trees.
Nonetheless, we considered this study to already be long and complex.
We therefore think that adding an extra question to this study (i.e. what is the effect of missing data on the morphological substitution rate parameter estimates?) might lose the reader and make an already long paper even longer.
Therefore, we added the results of these analyses to \textbf{Appendix-A Morphological rates estimations results}.
We link to this analysis in the methods as follows:

\textit{The detailed MrBayes parameters are available in Appendix A along with details on the $\alpha$ parameter estimation.} lines 336-338 @@@
See also Appendix A "Morphological rates estimations results".

@@

% -----------------
% Specific Comment 17
% -----------------
\item{\textcolor{blue}{\textbf{line 173:} streamline to "The substitution rates were selected from a gamma distribution with…" or something similar. }}

We modified the sentence as suggested.

% -----------------
% Specific Comment 18
% -----------------
\item{\textcolor{blue}{\textbf{line 173 to 176:} need some justification for why the substitution rate with alpha shape of 0.5 will avoid homoplasy. Yang 1996 is cited but clarify the justification of what is 'too much' homoplasy. }}

Since homoplasy is a difficult concept to measure (e.g. with character exhaustion; Wagner 2000, Evolution) it is difficult to assign a value for ``too much'' homoplasy.
We therefore removed this comment and changed the sentence to:
\textit{In practice, a value of $\alpha$ $<$ 1 decreases the number of sites with high substitution rates, thus reducing homoplasic sites and increasing the phylogenetic signal (Hassanin et al., 1998; Estoup et al., 2002).%\citep{Hassanin1998611,EstoupHomoplasy}.
Also, we chose this $\alpha$ value to be consistent with our protocol for simulating morphological characters (see below).} (line 189-192@@@).

@@

% -----------------
% Specific Comment 19
% -----------------
\item{\textcolor{blue}{\textbf{Line 176 to 180:} could be streamlined. It is not clear what is meant by 'no special assumption about how the characters evolved', what constitutes a special assumption? Could potentially be reduced to something like 'this model and these parameter settings strike a balance between realism for empirical datasets (citations) and parameter richness with more complex models (e.g., GTR, multiple partitions with independent models), making them more suitable for our computational limitations'. Just for example.}}

We modified the sentence as follows:

\textit{This model and these parameter settings strike a balance between realism for empirical datasets  (e.g. Douady, Kelly et al., 2014) and parameter richness with more complex models (e.g., GTR, multiple partitions with independent models), making them more suitable for our computational limitations (even with the parameters defined, the total computational time for the whole analysis was around 150 CPU years)}. lines @@@

@@

\begin{figure}[]
\centering
    \includegraphics[keepaspectratio=true]{response_fig1.pdf}
\caption{Figure 3 from Wright and Hillis 2014 with added x axis ticks. The red line represents the 0.5 rate.}
\end{figure}

We do agree, however, that if rates get lower than 0.5, variance in topological error increases (especially for the Mk model with missing data).
Therefore, we implemented the \textit{a posteriori} matrix selection step to our analysis (i.e. keeping only the ``complete'' matrices that led to trees with a median bootstrap $>$ 50; explained lines 259-268 in the submitted manuscript and lines @@@ and @@@ in the revised one).

We rephrased the section highlighted by the reviewer as follows:

\textit{We used low evolutionary rate parameters to be consistent with the molecular rate parameters, to avoid homoplasy in the morphological part of the matrix and create a clear phylogenetic signal (Wright and Hillis, 2014).
Topological error has been shown to be minimal at a morphological rate of 0.5 when using the M\textit{kv} model (Lewis, 2001; Wright and Hillis, 2014).
Note, however, that Wright and Hillis (2014) have shown that low morphological rates (< 0.5) increase variance in topological error, but we discarded simulations with such topological error by selecting only matrices with a “fair” phylogenetic signal (see Estimating phylogenies section below; Zander, 2004) so this should not influence our results.} lines 211-217 @@@

@@

% -----------------
% Specific Comment 23
% -----------------
\item{\textcolor{blue}{\textbf{line 201:} change 'get' for 'obtain' or something less colloquial.}}

We modified this whole section following reviewer 2 comment 2.

% -----------------
% Specific Comment 24
% -----------------
\item{\textcolor{blue}{\textbf{line 231-235:} streamline, e.g., 'The minimal dataset included in this study was 5\% of the morphological matrix for any taxon'.}}

We modified the sentence as follows: \textit{To avoid avoid matrices containing taxa without any data (morphological or molecular), we repeated the random deletion until the matrices contained at least 5\% of data for any taxon.} (line 264-267@@@).

@@

% -----------------
% Specific Comment 38
% -----------------
\item{\textcolor{blue}{\textbf{line 518 to 520:} I am not sure that (Wagner 2000) states that incongruence between molecular and morphological data is 'more important in small morphological matrices'. }}

We now refer to two more classical citations: Bremer and Struwe 1992 (American Journal of Botany) and Patterson et al. 1993 (Annual Review of Ecology and Systematics), on phylogenetic signal and congruence between morphological and molecular data and also refer to a classical example of conflict probably due to having a small matrix (Masters and Brothers 2002; American Journal of Physical Anthropology; 36 characters) lines 578-579@@@.

% -----------------
% Specific Comment 39
% -----------------
\item{\textcolor{blue}{\textbf{line 520:} Either change 'size' to plural or change 'were' to 'was', the former seems to flow better. }}

We modified the sentence as follows: \textit{The sizes of our data matrices were constrained by the performance of our protocol} (line 579-580@@@).

@@

We modified the sentence as suggested.

% -----------------
% Specific Comment 43
% -----------------
\item{\textcolor{blue}{\textbf{lines 542-547:} These seem like the most important practical results and should be more strongly emphasized. These are very important results for practitioners and the authors should try to make it more clear to the reader that this is a key take-home message.}}

We modified the paragraph to:

\textit{Surprisingly, the number of missing living taxa with morphological data ($M_{L}$) and the overall number of missing morphological characters ($N_{C}$), have a bigger effect than the amount of missing data for the fossil taxa ($M_{F}$).
For any additional missing living taxa with morphological data ($M_L$) beyond 50\%, there is no difference among trees with any combination of the other parameters ($M_F$ and $N_C$; Figure 6).
In other words, when the number of missing living taxa reaches 50\%, neither the amount of missing data in the fossil record ($M_F$), nor the number of characters in the matrix ($N_C$) affect topology.
A similar effect can be observed when the $N_C$ parameter reaches 50 characters (Figure 6).
This has important practical implications, especially for the best strategy to improve topology by collecting more morphological data (see below).} lines 589-609@@@

@@

% -----------------
% Specific Comment 46
% -----------------
\item{\textcolor{blue}{\textbf{line 549-565:} again, I am left wondering how much the differences in performance of max. likelihood and Bayesian inference techniques are driven by the use of the 'true' tree as a starting tree in the Bayesian MCMC search, while the likelihood analysis had no such guide tree to begin with. The authors at least need to make this point clear, if not re-run some subset of analyses without introducing a starting tree to test the effects of using the 'true' tree as a starting tree or not. I understand the computational limitations, but the practical limitation that empiricists will never have the 'true' tree to start with may (or may not) have a large impact on our interpretation and the generality of these findings. }}

As discussed above (reviewer 1, specific comment 29), we understand the reviewer's concern and have added an analysis in Appendix A (figure A2 and tables A1 and A2) where we tested the effect of using the ``true'' tree as a starting tree. We found that there is no significant effect of using the ``true'' tree instead of a random tree (MrBayes default) as a starting tree on topological recovery.

We refer to this supplementary analysis in the main text as follows:

\textit{Our results show that the topology of the Bayesian consensus tree is always closer to the ``best'' tree topology than the ``best'' Maximum Likelihood tree (Figure 5).
Note that the methodological choice of using the ``true'' tree as a starting tree for the Bayesian Inference rather than a random starting tree (see Methods), had no significant effect on topological recovery (see Appendix A, section ``Effect of the starting tree on Bayesian inference'' for details on the analysis of the effect of the starting tree in Bayesian inference).} lines 618-624@@@

@@

% -----------------
% Specific Comment 48
% -----------------
\item{\textcolor{blue}{\textbf{line 599:} remove extra close parenthesis inside parenthetical statement}}

We fixed this typo.

% -----------------
% Specific Comment 49
% -----------------
\item{\textcolor{blue}{\textbf{line 595-606:} this paragraph seems unsatisfactory advice for practitioners.
First, advising that the tree topology be fixed using the Bayesian consensus tree before conducting analyses such as dating strongly conditions the results on only a single tree and as the authors point out in the parenthetical statement, the dating information could improve accuracy. Indeed one of the many advantages of tip-dating is the joint inference of tree topology and divergence times.
The following statements that the posterior distribution should not be discarded does not leave the reader with a clear path that they should follow - if only the fixed tree topology is used to estimate divergence times, how can the substitution rate-scaled posterior distribution of highly probable trees be used for most comparative analyses which require time-scaled branch lengths?
This paragraph requires further justification and this study does have important bearing on a subject that gets little attention. }}

We revised the paragraph as follows:
\textit{Another practical implication of our results regards the tree inference methods.
Because the Bayesian consensus trees consistently recovered topologies closer to the ``best'' tree topology than the Maximum Likelihood trees, we advise that where a topological constraint is needed, Bayesian consensus trees should be used. This may apply to tree inferences using the Total Evidence method such as tip-dating (e.g. Ronquist et al., 2012a; Wood et al., 2013; Matzke, 2014). It is, however, possible that including dating information during tree inference could also improve the accuracy of the Bayesian posterior tree distribution, so a fixed topology should be used with caution.
Using the Bayesian consensus tree rather than the Maximum Likelihood can also reduce the number of false positive topologies (Swofford et al., 2001).
As shown in figure 5 and discussed in the section above (Effects of tree inference methods), the Bayesian consensus tree is more likely to not resolve poorly supported nodes due to missing data than the Maximum Likelihood tree that is more likely to incorrectly resolve such nodes (i.e. creating a false positive node).
Note, however, that we do not suggest discarding the Bayesian posterior tree distributions even though they performed poorly in recovering the ``best'' tree topology in our simulations (this can probably be imputed to the difficulties comparing distributions of trees; see above).
These trees will be invaluable for phylogenetic comparative analyses. For example a sub-sample of posterior trees distributions can be used to assess macroecological questions while better taking into account topological uncertainty (e.g. Fritz et al. 2009 and Jetz et al. 2012 trees used in Healy et al. 2014).} lines 680-700@@@

@@

We also linked these results to previous studies (Wiens 2003; Systematic Biology) and discussed the negative values of the Normalised Triplets metric in the reviewer 1 comment 10 (see above).

% -----------------
% Specific Comment 57
% -----------------
\item{\textcolor{blue}{\textbf{Figure 6:} I think this may be the most important figure because it is demonstrating the probability of overlap between the distributions of the best trees and the missing data trees.
It took me a long time to figure it out, however, and to figure out why the top of the triangle is all orange.
I would have expected something more like in Appendix C Figure C2 or Fig. C3.
At first I understood this to indicate high probability of overlap of the missing data trees which had the most missing data with the best tree.
But after re-reading the legend and methods several times, I am interpreting it as follows: the top of the triangle indicates a similarly poor fit of the distributions of missing data trees to the best tree under all parameter combinations once there is a high proportion of missing data.
The point is raised in the Results that once missing data in the living portion is > 50\% then they all perform poorly, but it was not easy to grasp.
Somehow this figure needs to be spelled out better in the caption and results so the reader can grasp it more quickly and easily without having to go back to the methods section to completely figure out what is being represented.}}

We thank the reviewer for pointing out this difficulty and acknowledge that such graphical representations of so many pairwise comparisons might be confusing.
We have made the figure caption clearer, and the methods section explaining how we calculated this figure and the results.

We modified the figure caption as follows:
\textit{The effects of missing data on topological recovery using Bayesian consensus trees. Both axes show the percentage of missing data from 0\% (white) to 75\% (black) for the three parameters: $M_{L}$ (upper line), $M_{F}$ (middle line) and $N_{C}$ (lower line). The topological recovery is measured as (A) the Normalised Robinson-Foulds metric and (B) the Normalised Triplets metric calculated using the Bhattacharyya Coefficient. The Bhattacharyya Coefficient values are indicated using a color gradient ranging from low probability of overlap in blue, to high probability of overlap in orange. Blue regions denote a poor overlap in Normalised metric values between the different parameter combinations (i.e. the parameters have a strong effect on the metric and thus the topological recovery). Conversely, orange regions denote a high overlap in Normalised metric values between the different parameter combinations (i.e. the parameters have a weak effect on the metric and thus the topological recovery).}

We modified the methods section as follows:
\textit{Because of the difficulties in representing so many pairwise comparisons in a meaningful way, we summarized these results as a heat map of Bhattacharyya Coefficients (see Figure 6).
In this type of figure, parameters that have similar effects on recovering the ``best'' topology (either positive or negative effects) will be denoted by similar colour patches in the heat map representation of these comparisons (see Figure 6).} lines 434-439@@@

We modified the results section as follows:
\textit{Using both Normalised Robinson-Foulds and Normalised Triplets metrics from the Bayesian consensus trees, the parameter combination with no missing data (i.e. $M_{L}$ = 0\%, $M_{F}$ = 0\%, $N_{C}$ = 100) is always the most dissimilar to all the other parameter combinations (thin deep blue line at the base of Figure 6).
The Normalised Robinson-Foulds metric (median Bhattacharrya coefficient = 0.79; blue regions in Figure 6A), however, displays more dissimilarities than the Normalised Triplets metric (median Bhattacharrya coefficient = 0.81; blue regions in Figure 6B).
The orange upper triangle in Figure 6A shows a high probability of overlap of the Normalised Robinson-Foulds metric for the trees with the $M_{L}$ parameter $\geq$ 50\% (Figure 6A).
Once $M_{L}$ $\geq$ 50\%, there is no additional effect of $M_{F}$ and $N_{C}$, regardless of the amount of missing data in these parameters (Figure 6A).
Likewise, once $N_{C}$ $<$ 50, there is no additional effect of $M_{L}$ and $M_{F}$ as denoted by the high probability of Normalised Robinson-Foulds metric overlap (horizontal orange stripes between the blue regions Figure 6A).
In Figure 5 for the Normalised Robinson-Foulds metric, this can be interpreted as the overlap between the distributions once $M_L$=50\%.} lines 480-497@@@

@@

% -----------------
% Comment 2
% -----------------
\item{\textcolor{blue}{\textbf{Better explanation of the missing data:} I found the explanation of the missing data strategy very difficult to follow. First, they oddly use the verb "to infer" for matrices; are they inferring matrices? Second, it's not clear anywhere that the authors actually removed the molecular data from the fossil taxa, though I assume they did. I hope the authors can re-write in simpler language ("explain like I'm 5") for easier comprehension.}}

### 2015-08-11 Manuscript/MPESubmission/Revision/response.tex

The lines 180 and 181 states "All the molecular information for fossil taxa was replaced by missing data ("?")" in the former manuscript. We changed the verb "infer" (see specific comment number 16 from reviewer 1).

We assume the reviewer is referring to section "2.2 Removing data". We therefore modified the whole section to improve its clarity and simplicity as follows:

\textit{To explore the effect of missing morphological data on topological recovery, we removed various amounts of the ``complete'' matrix to obtain matrices with missing morphological data.
Hereafter, we call these matrices with missing morphological data the ``missing-data'' matrices.
Note that the amount of molecular data remained constant throughout our simulations: 1000 molecular characters for living taxa and no molecular data for fossil taxa (see above).
We removed morphological data using three data incompleteness parameters:
\begin{enumerate}
\item{The proportion of missing living taxa ($M_L$).} This first missing-data parameter corresponds to the proportion of living taxa with no morphological data. It represents the number of living taxa that are present in the matrix but have only molecular data available. This reflects the fact that, because of the increasing ease of collecting molecular data, morphological data for living species is rarely collected (Guillerme and Cooper, 2015). Therefore, many living species will have only molecular data available. In practice, we removed all the morphological data from randomly chosen living taxa with five different proportions: 0\%, 10\%, 25\%, 50\% or 75\% of living taxa with no morphological data.
\item{The proportion of missing data in the fossil record ($M_F$).} This missing data parameter represents the completeness of the fossil record. Due to preservation biases, missing data for fossil taxa are common (Sansom and Wills, 2013). In practice, we randomly removed a proportion of data among the fossil taxa with five different proportions: 0\%, 10\%, 25\%, 50\% or 75\% of overall missing data for the fossil taxa.
%Added this part following comment 4.
Note that 50\% missing data for fossil taxa does not mean that each fossil is missing 50\% of its morphological data.
Instead this 50\% refers to missing fossil data across the whole matrix.
Some fossils may retain 100\% of their data and others may lose all their data at this parameter value (up to a maximum of 95\% missing data, see below). % NC: normally you'd say up to a maximum rather than up to a minimum. Is what I've written correct?
\item{The number of morphological characters for both living and fossil taxa ($N_C$).} This parameter is not a missing data parameter \textit{per se} but rather an indication of the size of the matrix. Any morphological matrix of any size has indeterminate missing data, given that the total number of characters is undefined, but presumably large. Therefore, this parameter corresponds to the overall number of characters available for both living and fossil taxa. In practice, we randomly removed whole characters from the morphological matrix reducing it to: 100, 90,75, 50 or 25 characters.
Note that these levels are equivalent to the two other parameters (i.e. 0\%, 10\%, 25\%, 50\% or 75\% of ``missing'' morphological characters).
\end{enumerate}}

\textit{Each parameter represents a different way of removing data from the morphological part of the matrix: $M_L$ removes entire rows from the living data; $M_F$ removes cells from the fossil data; and $N_C$ removes columns across both living and fossil data. Note that $M_L$  and $M_F$ differ not only because of the region of the matrix affected: for $M_L$ all the morphological data of a percentage of living taxa are removed, whereas for $M_F$ a percentage of the data are removed at random from across the whole of the morphological matrix for fossil taxa.}

\textit{We created matrices using all parameter combinations resulting in 125 ($5^3$) “missing-data” matrices. Note that one of these combinations ($M_L$=0\%; $M_F$=0\% and $N_C$=100) has no missing data so is equivalent to the “complete” matrix, thus we have one effectively complete matrix in our 125 “missing-data” matrices. In practice, we first removed the data following the two missing data parameters $M_L$ and $M_F$ and then removed data following the $N_C$ parameters. To avoid avoid matrices containing taxa without any data (morphological or molecular), we repeated the random deletion until the matrices contained at least 5\% of data for any taxon. Note that the living taxa always had at least 90\% of data (the 1000 molecular characters).} lines 222-267@@@

@@

In this study, we applied a similar approach to what the reviewer 2 suggests (i.e. no missing data for living taxa, and maximal number of characters but various amount of missing data for the fossil taxa). 
These results are present in figure 4, central row.
However, we think we may have not explained ourselves entirely clearly. The amount of changes among the fossil taxa was not set to any specific amount of missing data per fossil (e.g. like the "artificial fossil" approach in Pattinson et al 2014, Syst. Biol.) but we rather treated the missing data in the "whole" fossil record.
Because our fossil taxa and their characters where randomly simulated, we didn't have any \textit{a priori} assumption on how to remove the data.
We therefore decided to randomly remove a certain amount of data among all the fossils.
For example, when removing 75\% of the data for the fossil taxa ($M_F$=75\%) different scenarios could happen, ranging from a "homogeneous" incompleteness data distribution (each specific fossil taxa had 75\% of missing data) to a "heterogeneous" incompleteness data distribution  (80\% of the specific fossil taxa had 5\% available data and 21\% of the specific fossil taxa had 100\% available data).
In empirical matrices, the incompleteness data distribution probably lies in between both models.

We added the following comment to the explanation on removing data (See our response to reviewer 2's comment 1).

\textit{Note that 50\% missing data for fossil taxa does not mean that each fossil is missing 50\% of its morphological data.
Instead this 50\% refers to missing fossil data across the whole matrix.
Some fossils may retain 100\% of their data and others may lose all their data at this parameter value (up to a maximum of 95\% missing data, see below).}
